Fixing Vulnerabilities Hinders Maintainability

### 2019-02-05 paper/ICPC19/icpc19.tex
more fixes down

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines
or formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical evaluation;
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one of the
first steps taken by SIG to introduce security on their maintainability model).

BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small}
based on the idea that smaller codebases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.

@@

\setcounter{lstannotation}{0}
\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a \\Cross-Site Scripting vulnerability},label={lst:fix}]
 <ul>
  <li class='error'>
   <?php echo $l->t('Cloud not found');?>
   <br/>
   <p class='hint'>
-    <?php if(isset($_['file'])) echo $_['file']?>
   </p>
   <p class='hint'>
+    <?php if(isset($_['file'])) echo htmlentities($_['file'])?>
   </p> 
  </li>
 </ul>

\end{lstlisting} % \Luis{no listing lst:fix o que achas de fazer linebreak para
% cada uma das tags? assim nao havi o problema de ter a mesma linha dividida e ficava mais facil de ler:\\
% \texttt{<p class='hint'>}\\
% \texttt{~~<?php }\\
% \texttt{</p>}\\
% Also: reduz numero de espacos para cada tab -- 2 é suficiente.
% }
%
\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed in this section. The
methodology is comprised of the following steps, as illustrated in
Figure~\ref{fig:met}\Luis{como usamos numeros na enumeracao que descreve a
figura, esses numeros tb deveriam estar na figura.}:

@@

    \item \textbf{Miscellaneous (195 cases)} This pattern comprises several
    other categories of security refactorings (e.g., \textit{path traversal and
    buffer overflow})\Luis{exemplos ja nao precisa estar aqui porque sao
    enumerados logo a seguir}. It contains several other security refactorings
    that do not have their own pattern yet. The patterns not satisfying the
    size requirement of more than $20$ refactorings are the following ones:
    \textit{Resource Leak} (19), \textit{SHA-1 Hash Function} (1),
    \textit{Broken Access Control} (2), \textit{Sensitive Data Exposure} (18),
    \textit{Insufficient Attack Protection} (4), \textit{Overflow} (18),
    \textit{Path Traversal} (19).\Luis{Acho que este item pode ser reescrito de forma mais clara. sugestao:
    This category comprises the security refactorings that (1) do not relate to
    any particular pattern or (2) relate to patterns that do not occur in at
    least 20 refactorings: \textit{Resource Leak} (19), \dots} \end{itemize}

@@

One of the guidelines with which the project does not comply is the one presented in the report (cf. Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}. SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java). BCH evaluates this guideline based on the number 
\Luis{this seems unfinished. i think we ve already mention this Mccabe metric somewhere else.}

McCabe~\cite{1702388} is a software
metric that calculates the number of linearly independent paths of the source
code. BCH uses this metric to measure the number of branch points in the
\emph{Write Simple Units of Code} guideline.

The bar at the bottom of the figure
represents the top $30$ units that violate the guideline, sorted by severity,
The violation severities are indicated using colors, and there is a legend to
help to interpret the colors. The green bar represents the number of compliant
branch points per unit (\emph{at most $5$}), i.e., the units are compliant with
ISO 25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do
not comply with medium (\emph{above $5$}), high (\emph{above $10$}) and very high
(\emph{above $25$}) severity levels. In the bar, there are marks that pinpoint the
compliance thresholds for each severity level. If the green mark is somewhere
in the green bar it means it is compliant with a low level of severity.

@@

Aiming to analyze the impact of security refactorings, we use BCH to compute
maintainability of two different versions of the project (Figure~\ref{fig:commit}):
\begin{itemize}
	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
	refactoring;
	\item $v_{s}$, the version after fixing the security flaw, i.e., after the
	refactoring;
\end{itemize}

@@

\subsection{Statistical Validation}\label{sec:statsval}
%
To validate the maintainability differences in different groups of commits
(e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank
test with the significance level $\alpha = 0.05$~\cite{10.2307/3001968} . In
other words, we test the null hypothesis that the maintainability difference
between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a
security-commit) follows a symmetric distribution around $0$. However, when the
difference between groups is zero, the observations are discarded. Thus, we use
a more robust approach to the Wilcoxon test, the Wilcoxon test by Pratt which
provides an alternative that incorporates the cases where the maintainability
difference is zero~\cite{10.2307/2282543}. To understand the effect-size, as
advocated by the Common-language effect sizes\cite{graw:1992}, we compute the
mean difference, the median of the difference, and the percentage of cases that
reduce maintainability.
%
\section{Results}\label{sec:results}

@@

The impact of security and baseline commits in the maintainability of
open-source software is presented in Figure~\ref{fig:secvsreg}. For each type of
change, three types of impact of refactorings in the software maintainability
are presented: \emph{negative}, maintainability decreases (\emph{red});
\emph{none}, maintainability remains the same (\emph{yellow}); and,
\emph{positive}, maintainability increases (\emph{green}). Next to each
group, it is presented the mean ($\overline{x}$) and median (M)
of the maintainability difference and the p-value resulting from the Wilcoxon
signed-rank test. \Rui{this is not consistent to the statistical section...}

@@

Regarding security commits results, we observe that maintainability decreases in
$40.20\%$ ($244$), remains the same in $30.5\%$ ($185$) and increases in $29.33\%$
($178$) of the cases after security refactorings are applied. The resulting
p-value of the Wilcoxon signed-rank \Rui{consistentcy...} test is $1.21$x$10^{-10}$.
Since the p-value is below the significance level of $0.05$, we argue that
\emph{security refactorings have a negative impact in the maintainability of
open-source software}.

For regular commits, we observe that the maintainability decreases in $41.35\%$
($251$) of the cases after regular refactorings are applied. But in contrast
to security commits, it increases in $58.16\%$ ($353$) and remains the same in
$0.5\%$ ($3$) of the cases. The resulting p-value of the Wilcoxon signed-rank\Rui{consistency..}
test is $1.54$x$10^{-6}$. Since the p-value is below the significance level of
$0.05$, we argue that \emph{regular refactorings have a
positive impact in the maintainability of open-source software}.

@@

We have also evaluated the impact in the maintainability per programming language
of open-source software, Figure~\ref{fig:lang_main}. Although
the dataset contains security refactorings for more than $13$ languages, we only
present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP}
because one of the requirements of the Wilcoxon signed-rank test\Rui{consistency} is that the
input sample has to have more than $20$ samples. The other languages were
integrated into \emph{Others}. The results of the Wilcoxon signed-rank\Rui{consistency}
test by pattern indicate that there is statistical evidence to support that the
maintainability of open-source software decreases for security refactorings in
$3$ different languages: \emph{C} (p-value = $5.72$x$10^{-9}$), Ruby (p-value = $0.057$) and
PHP (p-value = $3.54$x$10^{-6}$). The results also provide statistical significance
for when security refactorings are performed in \emph{Java} (p-value =
$3.61$x$10^{-2}$). \emph{Python} and the languages that integrate the
\emph{Others} set contain more cases where the maintainability increases.
However, the results obtained for these patterns are not statistically
significant.

@@

The impact of each type of security refactoring on the maintainability of
open-source software is presented in Figure~\ref{fig:pat}. The results of the
Wilcoxon signed-rank test \Rui{again} per pattern indicated statistical
evidence supporting that the maintainability of open-source software decreases
for $4$ different patterns: \emph{Broken Autentication} (p-value = $0.018$),
\emph{Denial-of-Service}\Luis{em varios pontos deste paper este pattern era
referido com a sigla DoS. Por uma questao de consistencia com os outros
patterns sugiro que se escreva sempre por extenso (ou entao sempre com a
sigla).} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
\emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical
significance, results demonstrate evidence that maintainability is not
negatively impacted when removing \emph{Components with Known Vulnerabilities}
(p-value = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site Scripting
Vulnerabilities} (p-value = $0.001$). \emph{Injection} and \emph{Cross-Site
Request Forgery} produced more cases where the refactorings had a negative
impact in the maintainability. However, the results obtained for these patterns
are not statistically significant.

@@

\begin{figure}[h]
 	\centering
 	\includegraphics[width=0.57\textwidth]{figures/category.pdf}
 	\caption{Maintainability difference by pattern after security refactorings.
  \Luis{a median e' zero em muitos casos -- como interpretam isso? se nao houver
boa/simples explicacao talvez seja melhor usar o common language.}}
	\label{fig:pat}
\end{figure}

@@

\textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
%
We found that $40.2\%$ of the security refactoring have a negative impact on the
maintainability of open-source software (Figure~\ref{fig:secvsreg}). This raises
a new tradeoff when developers need to fix vulnerabilities on their projects
because they might be affecting the software maintainability or even introducing
new vulnerabilities. This study exhibits evidence that developers may have to
reduce maintainability for the sake of security. We understand that developers
should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	% \item \textbf{Security Frameworks} need to be secure and
	% user-friendly~\cite{5287006, 7676144}. Unfortunately, there are several
	% vulnerabilities that stay undisclosed. Thus, developers should not trust them
	% blindly and check defaults. \Rui{I don't understand the point.}

	\item \textbf{Documentation} of software should provide developers with the
	best practices to implement secure and maintainable software. Unmaintainable
	code is difficult to test, analyze and re-use.
	\item\textbf{Programming Languages} should provide new mechanisms to easily
	implement security patterns without endangering maintainability. \Rui{merge with
  3rd bullet?}
	Proper Authorization Mechanisms should be designed and
	implemented. This is one of the most complex security features to implement.
	Some mechanisms might assist authorization maintenance in long run, such as
	the combination of \texttt{HTTP} filters with \texttt{Spring Security}
	annotations. \Rui{not crystal clear -- do you mean programming languages
  should offer these mechanisms? What is that combination that you're talking
  about?}

	\item \textbf{Software Developers} should be avoided by developers. They should
	integrate tools such as BCH on their projects CI/CD pipelines to help them be
	more aware of the maintainability of their code and issues.

@@

\Rui{Intense, buzz --- this is not a blog. Don't understand the point very
well too.}

% Ray et al. (2017)~\cite{Ray:2017:LSP:3144574.3126905} and
% Berger et al. (2019)~\cite{2019arXiv190110220B}
%
% It would be expected that programming languages would have a more significant impact on the software maintainability according with previous work.
%
% According with the implications of previous work~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}, we would expect that the negative impact of programming languages on
% maintainability would be more significant
% surrounding the idea that poor design of programming languages and lack of using
% the best practices lead to more buggy and vulnerable
% code.

However,
Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of
more than $50\%$ on software maintainability. We suspect that these values are
the result of projects contributions policies. Nine out of ten projects --- the
equivalent to $31.8\%$ of the commits --- of the top-$10$ of dataset projects
with more contributors have restrict contribution policies regarding code
standards. Some examples are the \texttt{laravel/framework},
\texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
instance, \texttt{laravel/framework} requires the \texttt{PSR-2
Standard}\footnote{PSR-2 Standard available at
https://www.php-fig.org/psr/psr-2/ (Accessed on \today{})}.

@@

The impact of a refactoring depends on its complexity, i.e., if the refactoring
adds complexity to the codebase it is probably affecting the software
maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting}
and \emph{Using Components with Known Vulnerabilities} refactorings endure more
cases without any impact in the maintainability. Typically, these types of
vulnerabilities do not need extra lines to be fixed, as shown in
Listing~\ref{lst:fix}. Whereas, \emph{Broken Authentication} and vulnerabilities
such as the one presented in Section~\ref{sec:motivation} responsible by DoS
attacks are more difficult to fix (several code lines deleted and added).
Another example is the Apple authentication flaw discovered in $2017$
(CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user
could log in as root with an empty password. Patrick Wardle examined the cause
of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and
conclude that the flaw was due to the introduction of high cyclomatic complexity
in a method used to verify the password. In sum, \emph{Cross-Site Scripting} and
\emph{Using Components with Known Vulnerabilities} should take part of the
developers' preoccupations along with the other patterns that explicitly hinder
maintainability because despite their low complexity these are patterns that
still produce a considerable negative impact on the maintainability. Real
efforts have been made in creating and incorporating coding standards and best
practices in software production. However, security is still far from perfection
and fixing vulnerabilities might hinder software maintainability. Thus, it is
important to provide tool support to developers to help them apply these
patterns without harming software maintainability. It is also important to redesign programming languages. \Rui{and redesign programming
languages?}

@@

Many studies have investigated the relationship between refactorings and
software quality. Previous work focused on object-oriented metrics has evaluated the
impact of refactorings and exhibited proof that quantifying effectively the
impact of refactoring on maintainability may help to choose the appropriate
refactoring type~\cite{1167822}. In contrast to this work, Hegedus et
al.~\cite{HEGEDUS2018313} did not select particular metrics to assess the effect
of refactorings. Instead, statistical tests were used to find the metrics that
have the potential to change significantly after refactorings. The differences
of maintainability between groups of refactored elements and non-refactored were
analyzed. In addition, the authors conclude that the source code elements
subjected to refactorings had significantly lower maintainability than the
elements not affected. Studying the evolution of maintainability issues during
the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041}
discovered that maintainability decreases over time. Palomba et al.
($2018$)~\cite{Palomba:2018:DIM:3231288.3231337} exhibits proof that code smells
should be carefully monitored by programmers since there is a high correlation
between maintainability aspects and proneness to changes and faults. The present
work uses a model proposed by BCH and focuses solely on evaluating the impact of
patching vulnerabilities on software maintainability.

@@

Other studies investigated the effect of programming languages in software
quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. The first
study shows that some programming languages are more buggy than others, i.e.,
language design has an impact on software quality. However, the authors of the
second study were not able to reproduce the first one and did not obtain any
evidence about the impact of language design. Both have used the number of
defects as an indicator of software quality. Our works studies how
security refactorings affect software quality based on the code maintainability
analysis and provides evidence that programming language have impact on
maintainability.

@@

This study also shows evidence that there are security patterns that need more attention than
others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and
\emph{Memory Leaks} endure maintainability in $55.3\%$, $51.3\%$, and $47.5\%$,
respectively. \Sof{Check this par again}Following the implications of our work, we draw attention for the
need of code best practices documentation to
produce safe and maintainable code, proper authorization mechanisms design and
implementation, better tools to introduce in the CI/CD pipeline to help software
maintenance and new or improved programming languages to allow developers to
easily implement safety in their software without endangering maintainability.

>>>>

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines
or formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical evaluation;
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one of the
first steps taken by SIG to introduce security on their maintainability model). 
BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small}
based on the idea that smaller codebases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.

@@

\setcounter{lstannotation}{0}
\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a \\Cross-Site Scripting vulnerability},label={lst:fix}]
 <ul>
  <li class='error'>
   <?php echo $l->t('Cloud not found');?>
   <p class='hint'>
-   <?php if(isset($_['file'])) echo $_['file']?>
   </p>
   <p class='hint'>
+   <?php if(isset($_['file'])) echo htmlentities($_['file'])?>
   </p> 
  </li>
 </ul>

\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed in this section. The
methodology is comprised of the following steps, as illustrated in
Figure~\ref{fig:met}:

@@

    \item \textbf{Miscellaneous (195 cases)} This pattern comprises 
	several other security refactorings that do not have their own pattern yet and patterns that
	do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings:
    \textit{Resource Leak} (19), \textit{SHA-1 Hash Function} (1),
    \textit{Broken Access Control} (2), \textit{Sensitive Data Exposure} (18),
    \textit{Insufficient Attack Protection} (4), \textit{Overflow} (18),
    \textit{Path Traversal} (19). \end{itemize}

@@

SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed 
independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java). 
One of the guidelines with which the project does not comply is the one presented in 
the report (cf. Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}. BCH analyzes 
this guideline based on the McCabe Complexity~\cite{1702388} to calculate the number of branch 
points of a method. The bar at the bottom of the figure
represents the top $30$ units that violate the guideline, sorted by severity.
The violation severities are indicated using colors, and there is a legend to
help to interpret them. The green bar represents the number of compliant
branch points per unit (\emph{at most $5$}), i.e., the number of units are compliant with
ISO 25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do
not comply with medium (\emph{above $5$}), high (\emph{above $10$}) and very high
(\emph{above $25$}) severity levels. In the bar, there are marks that pinpoint the
compliance thresholds for each severity level. If the green mark is somewhere
in the green bar it means it is compliant with a low level of severity.

@@

Aiming to analyze the impact of security refactorings, we use BCH to compute
maintainability of two different versions of the project (cf. Figure~\ref{fig:commit}):
\begin{itemize}
	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
	refactoring;
	\item $v_{s}$, the version after fixing the security flaw, i.e., after the
	refactoring;
\end{itemize}

@@

\subsection{Statistical Validation}\label{sec:statsval}
%
To validate the maintainability differences in different groups of commits
(e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank
test with the significance level $\alpha = 0.05$~\cite{10.2307/3001968}. In
other words, we test the null hypothesis that the maintainability difference
between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a
security-commit) come from the same distribution. Nevertheless, this test has a 
limitation: it does not consider the groups of commits with a zero-difference 
maintainability. In $1959$, Pratt provided an improvement to the test to solve 
this issue making the test more robust. Thus, we use a version of the Wilcoxon 
test that incorporates the cases where maintainability is different from zero~\cite{10.2307/2282543}. 
To understand the effect-size, as
advocated by the Common-language effect sizes\cite{graw:1992}, we compute the
mean difference, the median of the difference, and the percentage of cases that
reduce maintainability.
%
\section{Results}\label{sec:results}

@@

The impact of security and baseline commits in the maintainability of
open-source software is presented in Figure~\ref{fig:secvsreg}. For each type of
change, three types of impact of refactorings in the software maintainability
are presented: \emph{negative}, maintainability decreases (\emph{red});
\emph{none}, maintainability remains the same (\emph{yellow}); and,
\emph{positive}, maintainability increases (\emph{green}). Next to each
group, it is presented the mean ($\overline{x}$) and median (M)
of the maintainability difference and the p-value resulting from the Paired Wilcoxon signed-rank test.

@@

Regarding security commits results, we observe that maintainability decreases in
$40.20\%$ ($244$), remains the same in $30.5\%$ ($185$) and increases in $29.33\%$
($178$) of the cases after security refactorings are applied. The resulting
p-value of the Paired Wilcoxon signed-rank test is $1.21$x$10^{-10}$.
Since the p-value is below the significance level of $0.05$, we argue that
\emph{security refactorings have a negative impact in the maintainability of
open-source software}.

For regular commits, we observe that the maintainability decreases in $41.35\%$
($251$) of the cases after regular refactorings are applied. But in contrast
to security commits, it increases in $58.16\%$ ($353$) and remains the same in
$0.5\%$ ($3$) of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
test is $1.54$x$10^{-6}$. Since the p-value is below the significance level of
$0.05$, we argue that \emph{regular refactorings have a
positive impact in the maintainability of open-source software}.

@@

We have also evaluated the impact in the maintainability per programming language
of open-source software, Figure~\ref{fig:lang_main}. Although
the dataset contains security refactorings for more than $13$ languages, we only
present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP}
because one of the requirements of the Paired Wilcoxon signed-rank is that the
input sample has to have more than $20$ samples. The other languages were
integrated into \emph{Others}. The results of the Paired Wilcoxon signed-rank
test by pattern indicate that there is statistical evidence to support that the
maintainability of open-source software decreases for security refactorings in
$3$ different languages: \emph{C} (p-value = $5.72$x$10^{-9}$), Ruby (p-value = $0.057$) and
PHP (p-value = $3.54$x$10^{-6}$). The results also provide statistical significance
for when security refactorings are performed in \emph{Java} (p-value =
$3.61$x$10^{-2}$). \emph{Python} and the languages that integrate the
\emph{Others} set contain more cases where the maintainability increases.
However, the results obtained for these patterns are not statistically
significant.

@@

The impact of each type of security refactoring on the maintainability of
open-source software is presented in Figure~\ref{fig:pat}. The results of the
Paired Wilcoxon signed-rank per pattern indicated statistical
evidence supporting that the maintainability of open-source software decreases
for $4$ different patterns: \emph{Broken Autentication} (p-value = $0.018$),
\emph{Denial-of-Service} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
\emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical
significance, results demonstrate evidence that maintainability is not
negatively impacted when removing \emph{Components with Known Vulnerabilities}
(p-value = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site Scripting
Vulnerabilities} (p-value = $0.001$). \emph{Injection} and \emph{Cross-Site
Request Forgery} produced more cases where the refactorings had a negative
impact in the maintainability. However, the results obtained for these patterns
are not statistically significant.

@@

\begin{figure}[h]
 	\centering
 	\includegraphics[width=0.57\textwidth]{figures/category.pdf}
 	\caption{Maintainability difference by pattern after security refactorings.
  \Luis{a median e' zero em muitos casos -- como interpretam isso? se nao houver
boa/simples explicacao talvez seja melhor usar o common language.}\Sof{Preciso de pensar melhor nisto}}
	\label{fig:pat}
\end{figure}

@@

\textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
%
We found that $40.2\%$ of the security refactoring have a negative impact on the
maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). This raises
a new tradeoff when developers need to fix vulnerabilities on their projects
because they might be affecting the software maintainability or even introducing
new vulnerabilities. This study exhibits evidence that developers may have to
reduce maintainability for the sake of security. We understand that developers
should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Documentation} of software must provide developers with the
	best practices to implement secure and maintainable software. Unmaintainable
	code is difficult to test, analyze and re-use.
	\item\textbf{Programming Languages} should provide new design patterns to easily
	implement security patterns without endangering maintainability. One example is 
	the need for designing new authorization mechanisms since these are one of the most 
	complex security features to implement.
	Some mechanisms might assist authorization maintenance in long run, such as
	the combination of \texttt{HTTP} filters with \texttt{Spring Security}
	annotations.

	\item \textbf{Software Developers} lack the knowledge to create secure and 
	maintainable code. Mainly because academic curriculums are not yet prepared 
	to educate these matters to them. Universities should do their part on making 
	this possible. Online services such as BCH have also a great potential to help 
	developers to be more aware of the maintainability issues introduced by their 
	changes. Thus, developers can put more work into improving their code maintainability 
	and avoid common issues such as code duplication.

@@

We would expect that the negative impact of programming languages on
maintainability would be more severe given the ideas that poor design of programming languages and lack of using
the best practices lead to more buggy and vulnerable
code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of
more than $50\%$ on software maintainability. We suspect that these values are
the result of projects contributions policies. Nine out of ten projects --- the
equivalent to $31.8\%$ of the commits --- of the top-$10$ of dataset projects
with more contributors have restrict contribution policies regarding code
standards. Some examples are the \texttt{laravel/framework},
\texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
instance, \texttt{laravel/framework} requires the \texttt{PSR-2
Standard}\footnote{PSR-2 Standard available at
https://www.php-fig.org/psr/psr-2/ (Accessed on \today{})}.

@@

The impact of a refactoring depends on its complexity, i.e., if the refactoring
adds complexity to the codebase it is probably affecting the software
maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting}
and \emph{Using Components with Known Vulnerabilities} refactorings endure more
cases without any impact in the maintainability. Typically, these types of
vulnerabilities do not need extra lines to be fixed, as shown in
Listing~\ref{lst:fix}. Whereas, \emph{Broken Authentication} and vulnerabilities
such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
attacks are more difficult to fix (several code lines deleted and added).
Another example is the Apple authentication flaw discovered in $2017$
(CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user
could log in as root with an empty password. Patrick Wardle examined the cause
of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and
conclude that the flaw was due to the introduction of high cyclomatic complexity
in a method used to verify the password. In sum, \emph{Cross-Site Scripting} and
\emph{Using Components with Known Vulnerabilities} should take part of the
developers' preoccupations along with the other patterns that explicitly hinder
maintainability because despite their low complexity these are patterns that
still produce a considerable negative impact on the maintainability. Real
efforts have been made in creating and incorporating coding standards and best
practices in software production. However, security is still far from perfection
and fixing vulnerabilities might hinder software maintainability. Thus, it is
not only important to provide tool support to developers to help them apply these
patterns without harming software maintainability, but also to create and redesign 
programming languages to facilitate developers improving software maintainability.

@@

Many studies have investigated the relationship between refactorings and
software quality. Previous work focused on object-oriented metrics has evaluated the
impact of refactorings and exhibited proof that quantifying effectively the
impact of refactorings on maintainability may help to choose the appropriate
refactoring type~\cite{1167822}. In contrast to this work, Hegedus et
al.~\cite{HEGEDUS2018313} did not select particular metrics to assess the effect
of refactorings. Instead, statistical tests were used to find the metrics that
have the potential to change significantly after refactorings. The differences
of maintainability between groups of refactored elements and non-refactored were
analyzed. In addition, the authors conclude that the source code elements
subjected to refactorings had significantly lower maintainability than the
elements not affected. Studying the evolution of maintainability issues during
the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041}
discovered that maintainability decreases over time. Palomba et al.
($2018$)~\cite{Palomba:2018:DIM:3231288.3231337} exhibits proof that code smells
should be carefully monitored by programmers since there is a high correlation
between maintainability aspects and proneness to changes and faults. The present
work uses a model proposed by BCH and focuses solely on evaluating the impact of
patching vulnerabilities on software maintainability.

@@

Other studies investigated the effect of programming languages in software
quality~\cite{Ray:2014:LSS:2635868.2635922,Ray:2017:LSP:3144574.3126905}. The first
study shows that some programming languages are more buggy than others, i.e.,
language design has an impact on software quality. However, the authors of the
second study were not able to reproduce the first one and did not obtain any
evidence about the impact of language design. Both have used the number of
defects as an indicator of software quality. In addition, 
Berger et al. ($2019$)~\cite{2019arXiv190110220B} tried to reproduce~\cite{Ray:2014:LSS:2635868.2635922, 
Ray:2017:LSP:3144574.3126905} and identified flaws that throw into distrust the previously demonstrated 
correlation between programming language and software defects. Our works studies how
security refactorings affect software quality based on the code maintainability
analysis and provides evidence that programming language have impact on
maintainability.

@@

This study also shows evidence that there are security patterns that need more attention than
others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and
\emph{Memory Leaks} endure maintainability in $55.3\%$, $51.3\%$, and $47.5\%$,
respectively. Following the implications of our work, we draw attention for the
need of better documentation explaining how to use best practices to produce secure 
and maintainable code; new and better programming languages; new academic curriculums 
in the fields of security and software quality; and, tools such as BCH to help developers 
predit the effect of their patches.
